---
title: "pisa_eda"
author: "Ian Drea Riley"
date: "03/13/2015"
output: html_document
---

```{r, echo=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
# load scales library to turn off scientific notation
library(scales)
df <- read.csv('pisa2012_working.csv')
```

```{r, echo=FALSE}
### UNIVARIATE ANALYSIS ###

qplot(math_score, data = df, binwidth = 20)
qplot(reading_score, data = df, binwidth = 20)
qplot(science_score, data = df, binwidth = 20)
```

Distributions for scores are normal
```{r, echo=FALSE}
qplot(country, data = df) +
  coord_flip()
```

Mass., Conn., FLA all listed as separate from United States. These should be
aggregated.
```{r, echo=FALSE}
qplot(tardy, data = df)
qplot(absent_day, data = df)
qplot(absent_class, data = df)
```

The absent_class variable is confusing - this seems to say that every student
missed at least one class, which seems wrong.
```{r, echo=FALSE}
qplot(wealth, data = df, binwidth = .5)
```

A strangely normal distribution for wealth. I'd expect it to be more right
skewed. I'm assuming this is a logarithmic scale.
```{r, echo=FALSE}
qplot(socio_economic_index, data = df)
```

Normalish, left skew. Range similar to wealth
```{r, echo=FALSE}
qplot(math_anxiety, data = df, binwidth = .25)
summary(df$math_anxiety)
```

Normal distribution with some hiccups.
```{r, echo=FALSE}
qplot(home_possessions, data = df)
```

Curious about scale used here. distributino normal with slight left skew
```{r, echo=FALSE}
qplot(gender, data = df)
```

Roughly even split. Slightly more women than men
```{r, echo=FALSE}
qplot(region, data = df) +
  coord_flip()
```

Will have to figure out how to translate these numbers into meaningful regions
```{r, echo=FALSE}
qplot(immegration_age, data = df, binwidth = 1)
```

Surprising that so few values are 0 - this would seem to suggest that the bulk
of students are immigrants, which seems unlikely.
```{r, echo=FALSE}
summary(df$immegration_age)
```

This may be a useless value, as there are so many NAs (98%). It's possible that
kids who were born in the country of the test simply did not fill this out
because it seemed not applicable (that is, they never immigrated, so even 0
is not a logical answer). That said, one can also immagine that there may be a
non-response bias here, where kids who are immigrants are reluctant to say so,
so imputing a '0' value for all NAs may be misguided.
```{r, echo=FALSE}
qplot(birth_country, data = df)
summary(df$birth_country)
```

A good, low number of NAs (~1.9%). This may help make a case for imputing values
for the immigration age variable.

## UNIVARIATE ANALYSIS

Many of the distributions are normal, which is nice. Consider a strategy for
imputing values across the immigration_age variable. Also, is the region code
actually necessary? It may not be unless I'm going to compare regions, which
may be very difficult if they're not in the US. Check whether the States that
have a different country value than the United States of America (Florida,
Connecticut and Massachusetts) are already included in the US value and can be
dropped, or if they should be aggregated into the US.

```{r, echo=FALSE}
### BIVARIATE ANALYSIS ###

qplot(math_score, reading_score, data = df, alpha = I(0.1))
qplot(math_score, science_score, data = df, alpha = I(0.1))
```

The scores from the tests have strong linear correlation. The variance for
science scores given a math score appears smaller than the variance for
reading scores given a math score. Given that Science is more math-based, this
is not surprising.
```{r, echo=FALSE}
qplot(wealth, math_score, data = df, alpha = I(0.01))
with(df, cor.test(wealth, math_score, data = df))
```

Interestingly, there does not seem to be a very strong correlation between
wealth and math scores.
```{r, echo=FALSE}
qplot(immegration_age, reading_score, data = df, alpha = I(0.1),
      geom = 'jitter')
qplot(immegration_age, science_score, data = df, alpha = I(0.1),
      geom = 'jitter')
qplot(immegration_age, math_score, data = df, alpha = I(0.1),
      geom = 'jitter')
with(df, cor.test(immegration_age, reading_score, data = df))
with(df, cor.test(immegration_age, math_score, data = df))
with(df, cor.test(immegration_age, science_score, data = df))
```

Immigration age seems to have no affect on scores
```{r, echo=FALSE}
qplot(birth_country, math_score, data = df, geom = "boxplot")
```

Country of birth appears to have no affect on math score, however the median
score for students who did not respond (NAs) is significantly lower than for
students who did. There is some non-response bias at work.
```{r, echo=FALSE}
qplot(math_anxiety, math_score, data = df, geom = "jitter", alpha = I(0.1))
with(df, cor.test(math_anxiety, math_score))
```

While the graph is certainly ugly, there is some correlation between math
anxiety and math score
```{r, echo=FALSE}
qplot(math_anxiety, reading_score, data = df, geom = "jitter", alpha = I(0.1))
with(df, cor.test(math_anxiety, reading_score))
```

There is also some correlation between math anxiety and reading score, however,
it is less than with math score.
```{r, echo=FALSE}
qplot(country, math_score, data = df, geom = "boxplot") +
  coord_flip()
```

Notable difference amongst countries
```{r, echo=FALSE}
qplot(tardy, math_score, data = df, geom = "boxplot")
qplot(absent_day, math_score, data = df, geom = "boxplot")
qplot(absent_class, math_score, data = df, alpha = I(0.01))
```

Lateness does affect classes. Also note non-response bias: NA box is similar to
Five or more times. The absent_class variable needs to be changed to an
ordinal variable.
```{r, echo=FALSE}
qplot(gender, math_score, data = df, geom = "boxplot")
qplot(gender, reading_score, data = df, geom = "boxplot")
qplot(gender, science_score, data = df, geom = "boxplot")
```

Interestingly, women do better with reading comprehension, men do slightly
better with math, and they are about the same with science.
```{r, echo=FALSE}
qplot(socio_economic_index, math_score, data = df, alpha = I(0.01))
qplot(socio_economic_index, reading_score, data = df, alpha = I(0.01))
qplot(socio_economic_index, science_score, data = df, alpha = I(0.01))
with(df, cor.test(socio_economic_index, math_score))
with(df, cor.test(socio_economic_index, reading_score))
with(df, cor.test(socio_economic_index, science_score))
```

Certainly a strong correlation between socio-economic index and math_score
```{r, echo=FALSE}
qplot(home_possessions, math_score, data = df)
```

Similar shape to wealth v. math_score graph
```{r, echo=FALSE}
qplot(wealth, data = df, binwidth = .5) +
  facet_wrap(~ country, scales = "free_y")
```

Significant differences in wealth distributions across countries.

### The following plots are using median values of country aggregates
```{r, echo=FALSE}
by_country_vars <- c("wealth", "socio_economic_index", "math_score",
                     "reading_score", "science_score", "math_anxiety")

#aggregate points by country and median value
country_agg <- df[by_country_vars]
country_agg <- aggregate(country_agg,
                         by = list(df$country),
                         FUN = median, na.rm = TRUE)
country_agg <- rename(country_agg, c(Group.1="country", wealth="wealth",
                                     math_score="math_score",
                                     reading_score="reading_score",
                                     science_score="science_score",
                                     math_anxiety="math_anxiety",
                                     socio_economic_index="escs"))

qplot(escs, math_score, data = country_agg)
qplot(wealth, math_score, data = country_agg)
with(country_agg, cor.test(escs, math_score))
with(country_agg, cor.test(wealth, math_score))
```

While wealth is a significantly worse predictor of performance than is the escs,
its median value by country proves a good predictor of the countries average
performance on math tests.