---
title: "pisa_eda"
author: "Ian Drea Riley"
date: "03/13/2015"
output: html_document
---

```{r, echo=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
# load scales library to turn off scientific notation
library(scales)
```

```{r, echo=FALSE}
### FUNCTIONS ###


#Creates a scatter plot from df$x and df$y with specified parameters
#Takes a data frame, as well as string values for the field names of x and y.
#Optionally takes x and y limits of the viewport, as well as x and y breaks, x
#and y jitter.
#Returns a plot object.
make_scatter <- function(df, x, y, xlims = NULL, ylims = NULL, xbrks = NULL,
                         ybrks = NULL, alpha = 1.0, jwidth = 0, jheight = 0) { 
  
  #create plot
  plt <- ggplot(aes_string(x = x, y = y), data = df) +
    geom_point(position = position_jitter(width = jwidth, height = jheight),
               alpha = alpha)
    
  
  #use optional parameters, if supplied
  if (!is.null(xlims)) {
    plt <- plt + coord_cartesian(xlim = xlims)
  }
  if (!is.null(ylims)) {
    plt <- plt + coord_cartesian(ylim = ylims)
  }
  if (!is.null(xbrks)) {
    plt <- plt + scale_x_continuous(breaks = xbrks)
  }
  if (!is.null(ybrks)) {
    plt <- plt + scale_y_continuous(breaks = ybrks)
  }
  
  #return plot
  plt
}

#Creates a simple bar plot. Ideally used for categorical variables.
#Takes a dataframe, as well as a string value for the field name of x.
#Optionally rotates the x axis 90 degrees
#Returns a plot object.
make_barplot <- function(df, x, rotate_x_lab = FALSE) {
  plt <- ggplot(aes_string(x = x), data = df) + geom_bar()
  if (rotate_x_lab) {
    plt <- rotate_labels(plt, x_angle = 90)
  }
  
  plt
}

#Rotates the labels of a plot by specified degrees.
#Takes a plot object.
#Optionally takes an angle of rotation for x and y labels separately.
#Returns a plot object.
rotate_labels <- function(plt, x_angle = 0, y_angle = 0) {
  rotated_plt = plt + theme(axis.text.x = element_text(angle = x_angle,
                                                       hjust = 1),
                            axis.text.y = element_text(angle = y_angle,
                                                       hjust = 1))
  rotated_plt
}

make_hist <- function(df, x, bwidth = NULL, breaks = NULL, limits = NULL) {
  plt <- ggplot(aes_string(x = x), data = df) +
    geom_histogram(binwidth = bwidth)
  
  if (!is.null(breaks)) {
    plt <- plt + scale_x_continuous(breaks = breaks)
  }
  if (!is.null(limits)) {
    plt <- plt + scale_x_continuous(limits = limits)
  }
  
  plt
}

get_nth_percentile <- function(field, percentile, remove_na = TRUE) {
  c(0, quantile(field, percentile, na.rm = remove_na))
```

```{r, echo=FALSE}
### UNIVARIATE ANALYSIS ###

qplot(math_score, data = df, binwidth = 20)
qplot(reading_score, data = df, binwidth = 20)
qplot(science_score, data = df, binwidth = 20)
```
Distributions for scores are normal
```{r, echo=FALSE}
qplot(country, data = df) +
  coord_flip()
```
Mass., Conn., FLA all listed as separate from United States. These should be
aggregated.
```{r, echo=FALSE}
qplot(tardy, data = df)
qplot(absent_day, data = df)
qplot(absent_class, data = df)
```
The absent_class variable is confusing - this seems to say that every student
missed at least one class, which seems wrong.
```{r, echo=FALSE}
qplot(wealth, data = df)
```
A strangely normal distribution for wealth. I'd expect it to be more right
skewed. I'm assuming this is a logarithmic scale.
```{r, echo=FALSE}
qplot(socio_economic_index, data = df)
```
Normalish, left skew. Range similar to wealth
```{r, echo=FALSE}
qplot(math_anxiety, data = df, binwidth = .25)
summary(df$math_anxiety)
```
Normal distribution with some hiccups.